The Ranked Nearest Neighbour Interchange space

### 2019-05-23 paper.tex
update proof lemma:caterpillar_dist=diameter (induction base convexity)

% Also mention the exact algortihm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
% Maybe it's worth trying the other algorithm in these trees as well.

@@

>>>>

% Also mention the exact algorithm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
% Maybe it's worth trying the other algorithm in these trees as well.

@@

\begin{lemma}
    Let $T$ and $T'$ be two caterpillar trees with distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
    There is a shortest path between $T$ and $T'$ that is a caterpillar path.
    \label{lemma:caterpillar_dist=diameter}
\end{lemma}

    % \todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
    %
    % \begin{figure}[H]
    %     \centering
    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
    %     \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
    %     \label{fig:caterpillar_max_distance}
    % \end{figure}

\begin{proof}
    Let $d_c(T,T')$ denote the length of a shortest caterpillar path from $T$ to $T'$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $T'$.
    We need to prove that $d(T,T') = \Delta(\rnniu)$ if, and only if, $d_c(T,T') = \Delta(\rnniu)$.

    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-2-k$ exchanges of taxa.
    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$, if the distance is $d(T,T') \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.

    Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ ad $T'$:
    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
    Otherwise, we re-label $T$ and $T'$.
    Let $T'$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.

    Induction basis: $n=3$\\
    All trees have distance $1$ and are caterpillar trees.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T$, $T'$ on $n$ taxa it is $d_c(T,T') = d(T, T') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu)$.
    Since we can reach $T'$ from $T$ by using Bubble Sort where each pair of taxa that is compared in that algorithm exchanges, we can conclude that taxon $n+1$ must be in the cherry of $T'$.
    It follows that there are on every path (not only caterpillar paths) from $T$ to $T'$ there are $n-1$ $\rnni$ moves needed to move the parent of $n+1$ down in the tree, since each $\rnni$ move can decrease the rank of an internal node by at most one.
    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the caterpillar path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
    Hence, the caterpillar distance $d_c(T|_n, T'|_n)$ is $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.

    In general, deleting taxon $n+1$ and its parent in every tree on any path $p$ from $T$ to $T'$ results in a path $p'$ from $T|_n$ to $T'|n$.
    Such a path $p'$ from $T|_n$ to $T'|n$ does not contain the $n-1$ $\rnni$ moves for moving the parent of $n+1$ down from the root to the cherry, so it is at least $n-1$ $\rnni$ moves shorter than $p$.
    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.

\end{proof}

\begin{theorem}
    The set of caterpillar trees is convex.
\end{theorem}

\begin{proof}
    We prove the Theorem by induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T_1$ and $T_2$.

    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T_1$, $T_2$ with distance $d(T_1,T_2) = \Delta(\rnniu)$ there is a caterpillar path from $T_1$ to $T_2$ between these trees.

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pait of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
    Let $T_1$ and $T_2$ be caterpillar trees with caterpillar distance $l$ \todo{define 'caterpillar distance'}.
    If $l \neq \Delta(\rnniu)$, there must be a pair of taxa $x,y$ that are neighboured in $T_2$ and have the same order in $T_1$ and $T_2$, which means that $rank(p_x) > rank(p_y)$ or $rank(p_x) < rank(p_y)$ holds in both trees $T_1$ and $T_2$.
    Then it follows that the tree $T_2'$ that equals $T_2$, but with exchanged taxa $x$ and $y$, has distance $l+1$ to $T_1$. \todo{really??}
    So there must be a shortest path from $T_2'$ to $T_1$ that only consists of caterpillar trees.
    Since it does not matter in which order the pairs of taxa that need to exchange their position do exchange on a path between two caterpillar trees, $T_2$ is on a shortest caterpillar path from $T_2'$ to $T_1$.
    Therefore, it must be $d(T_1, T_2) = l$.

\end{proof}
